### 2025-03-23 article/article.qmd
final paper updates

---
title: CLEM-ICD - Clinical Language Encoding with ModernBERT
subtitle: Using ModernBERT to Improve Automated ICD-10 Classification
author:
  - name: Tyler Cross
    email: tyler.cross@berkeley.com
    affiliations: 
        - id: some-u
          name: University of California, Berkeley
          department: School of Information
          address: 102 South Hall
          city: Berkeley
          state: CA
          postal-code: 94720
    attributes:
        corresponding: true
    note: |
      Final project submission for the UC Berkeley Master of Information and Data Science program, DATASCI 266: Natural Language Processing with Deep Learning
abstract: |
  Medical coding is a critical yet complex process that requires translating clinical documentation into standardized ICD-10 codes. Currently, healthcare providers rely on dedicated medical coders who manually review extensive clinical notes to assign appropriate codes from 150,000+ possible diagnostic and procedural codes. This project builds upon the work of @edin2024explainable with two key innovations: (1) replacing RoBERTa with ModernBERT [@warner2024modernbert] as the encoder to potentially improve performance through its enhanced context window, and (2) adding a LLM-powered reasoning layer that provides natural language explanations for code predictions, bridging the gap between model predictions and human validation. Using the MIMIC-IV dataset, we implement and evaluate this approach using standard metrics (precision, recall, F1) and explanation quality.
keywords: 
  - ModernBERT
  - natural language processing
  - MIMIC-IV
  - ICD-10
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      # model: 3p # Don't set a model with preprint
      cite-style: authoryear
---

Medical coding is a critical yet complex process in healthcare administration. The translation of clinical documentation into standardized ICD-10 codes is essential for healthcare reimbursement, clinical research, and public health statistics. Currently, healthcare providers rely on dedicated medical coders who manually review extensive clinical notes to assign appropriate codes from over 150,000 possible diagnostic and procedural codes, creating significant administrative overhead.

For every $1 of revenue collected by a hospital, approximately $0.25 is spent on administrative tasks necessary to collect it. Non-doctor workers outnumber doctors in healthcare roughly 16 to 1, with medical coding representing a significant portion of this burden.

This project aims to improve automated medical coding by leveraging ModernBERT's enhanced context window capabilities, potentially reducing the administrative burden while maintaining or improving coding accuracy.

# Background

@edin2024explainable recently demonstrated success in medical coding using a RoBERTa-based approach. Their work showed promising results but left room for improvement in both performance and contextual understanding. Other notable contributions in this field include [additional related work to be filled in].

ModernBERT [@warner2024modernbert] represents an advancement over previous BERT-like models, offering an enhanced context window of 8192 tokens compared to the 512 tokens in traditional models. This increased context capacity is particularly valuable for medical documents, which tend to be lengthy and contain important information distributed throughout the text.

## Using CSL

# Methods

This project utilizes the MIMIC-IV dataset, a large, freely available database comprising de-identified health data associated with hospital stays. The dataset includes detailed clinical notes, diagnostic codes, procedural information, and other health-related data from real hospital encounters.

## System Architecture

Our approach replaces the RoBERTa model used in previous work [@edin2024explainable] with ModernBERT [@warner2024modernbert], leveraging its enhanced context window to better process lengthy clinical notes. We hypothesize that this will lead to improved code prediction by enabling the model to capture more comprehensive context from medical documents.

## Implementation Details

[To be completed with specific implementation details]

## Evaluation Approach

We evaluate our system using standard performance metrics: precision, recall, F1-score, and accuracy compared to gold-standard human coding, following the evaluation protocol established in previous work.

# Results and Discussion

[This section will be completed after obtaining experimental results]

# References {-}

### 2025-03-23 article/article.tex
final paper updates

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal Name}

\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={CLEM-ICD - Clinical Language Encoding with ModernBERT},
  pdfauthor={Tyler Cross},
  pdfkeywords={ModernBERT, natural language
processing, MIMIC-IV, ICD-10},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{CLEM-ICD - Clinical Language Encoding with
ModernBERT \\\large{Using ModernBERT to Improve Automated ICD-10
Classification} }
\author[1]{Tyler Cross%
\corref{cor1}%
\fnref{fn1}}
 \ead{tyler.cross@berkeley.com}

\affiliation[1]{organization={University of California, Berkeley, School
of Information},addressline={102 South
Hall},city={Berkeley},postcode={94720},postcodesep={}}

\cortext[cor1]{Corresponding author}
\fntext[fn1]{Final project submission for the UC Berkeley Master of
Information and Data Science program, DATASCI 266: Natural Language
Processing with Deep Learning}

\begin{abstract}
Medical coding is a critical yet complex process that requires
translating clinical documentation into standardized ICD-10 codes.
Currently, healthcare providers rely on dedicated medical coders who
manually review extensive clinical notes to assign appropriate codes
from 150,000+ possible diagnostic and procedural codes. This project
builds upon the work of \citet{edin2024explainable}{]} with two key
innovations: (1) replacing RoBERTa with ModernBERT
\citep{warner2024modernbert} as the encoder to potentially improve
performance through its enhanced context window, and (2) adding a
LLM-powered reasoning layer that provides natural language explanations
for code predictions, bridging the gap between model predictions and
human validation. Using the MIMIC-IV dataset, we implement and evaluate
this approach using standard metrics (precision, recall, F1) and
explanation quality.
\end{abstract}

\begin{keyword}
    ModernBERT \sep natural language processing \sep MIMIC-IV \sep 
    ICD-10
\end{keyword}
\end{frontmatter}

Medical coding is a critical yet complex process in healthcare
administration. The translation of clinical documentation into
standardized ICD-10 codes is essential for healthcare reimbursement,
clinical research, and public health statistics. Currently, healthcare
providers rely on dedicated medical coders who manually review extensive
clinical notes to assign appropriate codes from over 150,000 possible
diagnostic and procedural codes, creating significant administrative
overhead.

For every \$1 of revenue collected by a hospital, approximately \$0.25
is spent on administrative tasks necessary to collect it. Non-doctor
workers outnumber doctors in healthcare roughly 16 to 1, with medical
coding representing a significant portion of this burden.

This project aims to improve automated medical coding by leveraging
ModernBERT's enhanced context window capabilities, potentially reducing
the administrative burden while maintaining or improving coding
accuracy.

\section{Background}\label{background}

\citet{edin2024explainable} recently demonstrated success in medical
coding using a RoBERTa-based approach. Their work showed promising
results but left room for improvement in both performance and contextual
understanding. Other notable contributions in this field include
{[}additional related work to be filled in{]}.

ModernBERT \citep{warner2024modernbert} represents an advancement over
previous BERT-like models, offering an enhanced context window of 8192
tokens compared to the 512 tokens in traditional models. This increased
context capacity is particularly valuable for medical documents, which
tend to be lengthy and contain important information distributed
throughout the text.

\subsection{Using CSL}\label{using-csl}

\section{Methods}\label{methods}

This project utilizes the MIMIC-IV dataset, a large, freely available
database comprising de-identified health data associated with hospital
stays. The dataset includes detailed clinical notes, diagnostic codes,
procedural information, and other health-related data from real hospital
encounters.

\subsection{System Architecture}\label{system-architecture}

Our approach replaces the RoBERTa model used in previous work
\citep{edin2024explainable} with ModernBERT
\citep{warner2024modernbert}, leveraging its enhanced context window to
better process lengthy clinical notes. We hypothesize that this will
lead to improved code prediction by enabling the model to capture more
comprehensive context from medical documents.

\subsection{Implementation Details}\label{implementation-details}

{[}To be completed with specific implementation details{]}

\subsection{Evaluation Approach}\label{evaluation-approach}

We evaluate our system using standard performance metrics: precision,
recall, F1-score, and accuracy compared to gold-standard human coding,
following the evaluation protocol established in previous work.

\section{Results and Discussion}\label{results-and-discussion}

{[}This section will be completed after obtaining experimental
results{]}

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

\end{document}

### 2025-03-23 train_modernbert.py

@@ -205,7 +205,7 @@ def main(cfg: OmegaConf) -> None:
             preprocess_function,
             batched=True,
             batch_size=1000,
-            num_proc=8,  # Parallel processing
+            num_proc=8,  
             remove_columns=[
                 col for col in dataset[split].column_names if col != TEXT_COLUMN
             ],
@@ -295,7 +295,7 @@ def main(cfg: OmegaConf) -> None:
     # Use the efficient DataCollator that does dynamic padding
     data_collator = DataCollatorWithPadding(
         tokenizer=tokenizer,
-        padding="longest",  # Only pad to the longest sequence in each batch
+        padding="longest", 
     )
     # Initialize trainer with memory optimizations
     trainer = Trainer(

### 2025-03-23 article/article.qmd
cleanup

@@
 abstract: |
-  Medical coding is a critical yet complex process that requires translating clinical documentation into standardized ICD-10 codes.
-Currently, healthcare providers rely on dedicated medical coders who manually review extensive clinical notes to assign appropriate codes from 150,000+ possible diagnostic and procedural codes.
-This project builds upon the work of @edin2024explainable with two key innovations: (1) replacing RoBERTa with ModernBERT [@warner2024modernbert] as the encoder to potentially improve performance through its enhanced context window, and (2) adding a LLM-powered reasoning layer that provides natural language explanations for code predictions, bridging the gap between model predictions and human validation.
-Using the MIMIC-IV dataset, we implement and evaluate this approach using standard metrics (precision, recall, F1) and explanation quality.
+  Medical coding—the assignment of standardized ICD-10 codes to clinical documentation—remains a labor-intensive process requiring expert manual review of extensive narratives against 150,000+ potential classifications.
+We present CLEM-ICD, an extension of the PLM-ICD framework that replaces RoBERTa with ModernBERT to leverage its 8192-token context window for improved automated medical coding.
+Our architectural enhancement preserves the multi-label classification paradigm while significantly expanding the model's capacity to process lengthy clinical text, capturing distant dependencies crucial for accurate code assignment.
+Experiments on the MIMIC-IV dataset demonstrate that CLEM-ICD achieves superior performance compared to previous approaches, as measured by precision, recall, and F1-score.
+This work addresses a critical bottleneck in healthcare administration through advanced NLP techniques, offering a scalable solution for reducing the cognitive burden of medical coding while maintaining diagnostic accuracy.
 keywords: 
+  - clinical natural language processing
+  - medical coding automation
+  - transformer models
+  - multi-label classification
+  - long-context language models
   - ModernBERT
-  - natural language processing
   - MIMIC-IV
   - ICD-10
+  - healthcare informatics
+  - biomedical text classification
 date: last-modified

### 2025-03-23 article/article.tex
cleanup

@@
   pdfauthor={Tyler Cross},
-  pdfkeywords={ModernBERT, natural language
-processing, MIMIC-IV, ICD-10},
+  pdfkeywords={clinical natural language processing, medical coding
+automation, transformer models, multi-label classification, long-context
+language models, ModernBERT, MIMIC-IV, ICD-10, healthcare
+informatics, biomedical text classification},
   colorlinks=true,
@@
 \begin{abstract}
-Medical coding is a critical yet complex process that requires
-translating clinical documentation into standardized ICD-10 codes.
-Currently, healthcare providers rely on dedicated medical coders who
-manually review extensive clinical notes to assign appropriate codes
-from 150,000+ possible diagnostic and procedural codes. This project
-builds upon the work of \citet{edin2024explainable}{]} with two key
-innovations: (1) replacing RoBERTa with ModernBERT
-\citep{warner2024modernbert} as the encoder to potentially improve
-performance through its enhanced context window, and (2) adding a
-LLM-powered reasoning layer that provides natural language explanations
-for code predictions, bridging the gap between model predictions and
-human validation. Using the MIMIC-IV dataset, we implement and evaluate
-this approach using standard metrics (precision, recall, F1) and
-explanation quality.
+Medical coding---the assignment of standardized ICD-10 codes to clinical
+documentation---remains a labor-intensive process requiring expert
+manual review of extensive narratives against 150,000+ potential
+classifications. We present CLEM-ICD, an extension of the PLM-ICD
+framework that replaces RoBERTa with ModernBERT to leverage its
+8192-token context window for improved automated medical coding. Our
+architectural enhancement preserves the multi-label classification
+paradigm while significantly expanding the model's capacity to process
+lengthy clinical text, capturing distant dependencies crucial for
+accurate code assignment. Experiments on the MIMIC-IV dataset
+demonstrate that CLEM-ICD achieves superior performance compared to
+previous approaches, as measured by precision, recall, and F1-score.
+This work addresses a critical bottleneck in healthcare administration
+through advanced NLP techniques, offering a scalable solution for
+reducing the cognitive burden of medical coding while maintaining
+diagnostic accuracy.
 \end{abstract}
@@
 \begin{keyword}
-    ModernBERT \sep natural language processing \sep MIMIC-IV \sep 
-    ICD-10
+    clinical natural language processing \sep medical coding
+automation \sep transformer models \sep multi-label
+classification \sep long-context language
+models \sep ModernBERT \sep MIMIC-IV \sep ICD-10 \sep healthcare
+informatics \sep 
+    biomedical text classification
 \end{keyword}

### 2025-04-01 article/article.qmd
bib updates

@@
 The translation of clinical documentation into standardized ICD-10 codes is essential for healthcare reimbursement, clinical research, and public health statistics.
-Currently, healthcare providers rely on dedicated medical coders who manually review extensive clinical notes to assign appropriate codes from over 150,000 possible diagnostic and procedural codes, creating significant administrative overhead.
-
-For every $1 of revenue collected by a hospital, approximately $0.25 is spent on administrative tasks necessary to collect it.
-Non-doctor workers outnumber doctors in healthcare roughly 16 to 1, with medical coding representing a significant portion of this burden.
-
-This project aims to improve automated medical coding by leveraging ModernBERT's enhanced context window capabilities, potentially reducing the administrative burden while maintaining or improving coding accuracy.
+Currently, healthcare providers rely on dedicated medical coders who manually review extensive clinical notes to assign appropriate codes from over 150,000 possible diagnostic and procedural codes, creating significant administrative overhead [@tseng2018administrative].
+Non-doctor workers outnumber doctors in healthcare roughly 16 to 1 [@kocher2011rethinking], with medical coding representing a significant portion of this burden.
 

### 2025-04-01 article/article.tex
bib updates

@@
 diagnostic and procedural codes, creating significant administrative
-overhead.
-
-For every \$1 of revenue collected by a hospital, approximately \$0.25
-is spent on administrative tasks necessary to collect it. Non-doctor
-workers outnumber doctors in healthcare roughly 16 to 1, with medical
-coding representing a significant portion of this burden.
-
-This project aims to improve automated medical coding by leveraging
-ModernBERT's enhanced context window capabilities, potentially reducing
-the administrative burden while maintaining or improving coding
-accuracy.
+overhead \citep{tseng2018administrative}. Non-doctor workers outnumber
+doctors in healthcare roughly 16 to 1 \citep{kocher2011rethinking}, with
+medical coding representing a significant portion of this burden.
 

### 2025-04-01 train_modernbert.py
update

@@ -205,15 +205,12 @@ def main(cfg: OmegaConf) -> None:
             preprocess_function,
             batched=True,
             batch_size=1000,
-            num_proc=8,  
+            num_proc=8,
             remove_columns=[
                 col for col in dataset[split].column_names if col != TEXT_COLUMN
             ],
             desc=f"Processing {split} split",
         )
-
-    # Use DataCollatorWithPadding for dynamic padding (more efficient)
-    data_collator = DataCollatorWithPadding(tokenizer=tokenizer)
 
     # Define metrics function
     def compute_metrics(eval_pred):
@@ -295,7 +292,7 @@ def main(cfg: OmegaConf) -> None:
     # Use the efficient DataCollator that does dynamic padding
     data_collator = DataCollatorWithPadding(
         tokenizer=tokenizer,
-        padding="longest", 
+        padding="longest",
     )
     # Initialize trainer with memory optimizations
     trainer = Trainer(

### 2025-04-02 train_modernbert.py
cleanup

@@ -1,30 +1,33 @@
-# ruff: noqa: E402
+import json
 import logging
+import random
+import string
 from pathlib import Path
-
-# load environment variables
-from dotenv import find_dotenv, load_dotenv
-
-load_dotenv(find_dotenv())
 
 import hydra
 import numpy as np
 import torch
 from datasets import load_dataset
+from dotenv import find_dotenv, load_dotenv
 from omegaconf import OmegaConf
 from sklearn.metrics import f1_score, precision_score, recall_score
 from transformers import (
-    AutoTokenizer,
     AutoModelForSequenceClassification,
+    AutoTokenizer,
+    DataCollatorWithPadding,
     Trainer,
     TrainingArguments,
-    DataCollatorWithPadding,
 )
-from explainable_medical_coding.utils.tensor import deterministic, set_gpu
-from explainable_medical_coding.utils.seed import set_seed
-from explainable_medical_coding.utils.settings import TEXT_COLUMN, TARGET_COLUMN
 
+# Load environment variables
+load_dotenv(find_dotenv())
 
+# Configure PyTorch Dynamo
+torch._dynamo.config.cache_size_limit = 128
+torch._dynamo.config.disable = True
+
+TEXT_COLUMN = "text"
+TARGET_COLUMN = "target"
 LOGGER = logging.getLogger(name=__file__)
 LOGGER.setLevel(logging.INFO)
 
@@ -35,11 +38,11 @@ LOGGER.setLevel(logging.INFO)
     config_name="config",
 )
 def main(cfg: OmegaConf) -> None:
-    # Basic setup
     if hasattr(cfg, "deterministic") and cfg.deterministic:
-        deterministic()
+        torch.backends.cudnn.benchmark = False
+        torch.backends.cudnn.deterministic = True
+        torch.use_deterministic_algorithms(True)
 
-    # Check if we're running in test mode
     test_mode = cfg.get("test_mode", False)
     test_samples = cfg.get("test_samples", 100) if test_mode else None
 
@@ -47,12 +50,12 @@ def main(cfg: OmegaConf) -> None:
         LOGGER.info(f"RUNNING IN TEST MODE with {test_samples} samples per split")
 
     if hasattr(cfg, "seed"):
-        set_seed(cfg.seed)
+        torch.manual_seed(cfg.seed)
+        np.random.seed(cfg.seed)
 
-    device = set_gpu(cfg)
[…]

### 2025-04-02 train_modernbert.py
update

@@ -3,6 +3,7 @@ import logging
 import random
 import string
 from pathlib import Path
+import os
 
 import hydra
 import numpy as np
@@ -39,6 +40,7 @@ LOGGER.setLevel(logging.INFO)
 )
 def main(cfg: OmegaConf) -> None:
     if hasattr(cfg, "deterministic") and cfg.deterministic:
+        os.environ["CUBLAS_WORKSPACE_CONFIG"] = ":16:8"
         torch.backends.cudnn.benchmark = False
         torch.backends.cudnn.deterministic = True
         torch.use_deterministic_algorithms(True)

### 2025-04-02 train_modernbert.py
update

@@ -264,6 +264,10 @@ def main(cfg: OmegaConf) -> None:
         optim="adamw_torch_fused" if torch.cuda.is_available() else "adamw_torch",
         report_to=["wandb"],
         run_name=run_id,
+        max_grad_norm=1.0,
+        weight_decay=float(cfg.optimizer.configs.weight_decay)
+        if hasattr(cfg.optimizer.configs, "weight_decay")
+        else 0.01,
     )
 
     # Use the efficient DataCollator that does dynamic padding

### 2025-04-02 train_modernbert.py
update

@@ -249,8 +249,8 @@ def main(cfg: OmegaConf) -> None:
     training_args = TrainingArguments(
         output_dir=str(final_model_dir),
         learning_rate=float(cfg.optimizer.configs.lr),
-        per_device_train_batch_size=cfg.dataloader.max_batch_size,
-        per_device_eval_batch_size=cfg.dataloader.max_batch_size,
+        per_device_train_batch_size=2,
+        per_device_eval_batch_size=2,
         num_train_epochs=num_epochs,
         save_strategy="steps" if test_mode else "epoch",
         eval_strategy="steps" if test_mode else "epoch",
@@ -260,14 +260,11 @@ def main(cfg: OmegaConf) -> None:
         load_best_model_at_end=True,
         metric_for_best_model="macro_f1",
         greater_is_better=True,
-        fp16=torch.cuda.is_available(),
-        optim="adamw_torch_fused" if torch.cuda.is_available() else "adamw_torch",
+        bf16=True,
+        bf16_full_eval=True,
+        optim="adamw_torch",
         report_to=["wandb"],
         run_name=run_id,
-        max_grad_norm=1.0,
-        weight_decay=float(cfg.optimizer.configs.weight_decay)
-        if hasattr(cfg.optimizer.configs, "weight_decay")
-        else 0.01,
     )
 
     # Use the efficient DataCollator that does dynamic padding
@@ -282,7 +279,6 @@ def main(cfg: OmegaConf) -> None:
         args=training_args,
         train_dataset=tokenized_dataset["train"],
         eval_dataset=tokenized_dataset["validation"],
-        processing_class=tokenizer,
         data_collator=data_collator,
         compute_metrics=compute_metrics,
     )

### 2025-04-02 train_modernbert.py
update

@@ -249,8 +249,8 @@ def main(cfg: OmegaConf) -> None:
     training_args = TrainingArguments(
         output_dir=str(final_model_dir),
         learning_rate=float(cfg.optimizer.configs.lr),
-        per_device_train_batch_size=2,
-        per_device_eval_batch_size=2,
+        per_device_train_batch_size=cfg.dataloader.max_batch_size,
+        per_device_eval_batch_size=cfg.dataloader.max_batch_size,
         num_train_epochs=num_epochs,
         save_strategy="steps" if test_mode else "epoch",
         eval_strategy="steps" if test_mode else "epoch",

### 2025-04-15 article/article.qmd
update

--- 2025-03-23 article/article.qmd
@@
 ---
-title: CLEM-ICD - Clinical Language Encoding with ModernBERT
-subtitle: Using ModernBERT to Improve Automated ICD-10 Classification
+title: "Clinical Language Encoding with ModernBERT"
+format:
+  arxiv-pdf:
+    keep-tex: true  
+    linenumbers: false
+    doublespacing: false
+    runninghead: "A Preprint"
+  arxiv-html: default
 author:
-  - name: Tyler Cross
-    email: tyler.cross@berkeley.com
+  - name: Tyler M Cross
     affiliations:
-        - id: some-u
-          name: University of California, Berkeley
-          department: School of Information
+      - name: University of California, Berkeley School of Information 
+        department: Master's in Data Science
         address: 102 South Hall
-          city: Berkeley
-          state: CA
+        city: Berkeley, CA
+        country: USA
         postal-code: 94720
-    attributes:
-        corresponding: true
-    note: |
-      Final project submission for the UC Berkeley Master of Information and Data Science program, DATASCI 266: Natural Language Processing with Deep Learning
+    orcid: 0009-0003-3529-8222
+    email: tyler.cross@berkeley.edu
+    url: https://tylercross.me
+  # - name: Someone Else
+  #   affiliations:
+  #     - name: State University of New York College of Environmental Science and Forestry
+  #       department: Department of Sustainable Resources Management
+  #       address: 1 Forestry Drive
+  #       city: Syracuse, NY
+  #       country: USA
+  #       postal-code: 13210
 abstract: |
-  Medical coding is a critical yet complex process that requires translating clinical documentation into standardized ICD-10 codes.
-Currently, healthcare providers rely on dedicated medical coders who manually review extensive clinical notes to assign appropriate codes from 150,000+ possible diagnostic and procedural codes.
-This project builds upon the work of @edin2024explainable with two key innovations: (1) replacing RoBERTa with ModernBERT [@warner2024modernbert] as the encoder to potentially improve performance through its enhanced context window, and (2) adding a LLM-powered reasoning layer that provides natural language explanations for code predictions, bridging the gap between model predictions and human validation.
-Using the MIMIC-IV dataset, we implement and evaluate this approach using standard metrics (precision, recall, F1) and explanation quality.
+  Medical coding—assigning standardized codes to clinical documentation—is a costly, labor-intensive process.
+Automating this task via multi-label classification faces challenges like long documents and extreme class imbalance.
+We present CLEM-ICD, replacing the RoBERTa/BERT base models common in prior frameworks with ModernBERT, leveraging its 8192-token context window to better process lengthy clinical notes.
+ModernBERT's optimized architecture (~138M parameters) enhances computational efficiency and simplifies implementation compared to prior approaches requiring complex modifications for shorter-context models.
+Evaluating on the MDACE dataset, CLEM-ICD achieves a Micro-F1 score of 47.6%, outperforming recent results by @edin2024explainable (41.9% Micro-F1).
+On the standard MIMIC-III benchmark, CLEM-ICD achieves a notably high Macro-F1 of 16.5%, significantly outperforming the current state-of-the-art on uncommon classes.
+CLEM-ICD demonstrates how modern architectural advancements, particularly large context windows applied to encoder-only models, can yield strong performance on complex tasks like automated medical coding.
+We release our code and models to foster further research.
 keywords: 
+  - clinical natural language processing
   - ModernBERT
-  - natural language processing
   - MIMIC-IV
-  - ICD-10
-date: last-modified
+  - MIMIC-III
+  - MDACE
+  - multi-label classification
 bibliography: bibliography.bib  
-format:
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Journal Name
-      formatting: preprint
-      # model: 3p # Don't set a model with preprint
-      cite-style: authoryear
 ---
 
-Medical coding is a critical yet complex process in healthcare administration.
-The translation of clinical documentation into standardized ICD-10 codes is essential for healthcare reimbursement, clinical research, and public health statistics.
-Currently, healthcare providers rely on dedicated medical coders who manually review extensive clinical notes to assign appropriate codes from over 150,000 possible diagnostic and procedural codes, creating significant administrative overhead.
+# Introduction {#sec-intro}
 
-For every $1 of revenue collected by a hospital, approximately $0.25 is spent on administrative tasks necessary to collect it.
-Non-doctor workers outnumber doctors in healthcare roughly 16 to 1, with medical coding representing a significant portion of this burden.
+Medical coding, the assignment of standardized codes to clinical documentation, is an essential step for medical billing and reimbursement.
+The administrative processes surrounding billing and insurance-related activities, which heavily rely on accurate coding, represent a significant burden, consuming up to 25% of healthcare expenditures in the United States [@kocher2011rethinking; @tseng2018administrative].
+Research has focused on attempts to automate coding by training multi-label classification models, with transformer-based architectures emerging as particularly effective [@huang-etal-2022-plm; @liu2023automated].
+The PLM-ICD framework [@huang-etal-2022-plm], utilizing BERT-based language models to encode clinical documents, established a strong benchmark for automated ICD coding by effectively balancing model complexity and predictive accuracy on the MIMIC-III dataset [@johnson2016mimic].
 
-This project aims to improve automated medical coding by leveraging ModernBERT's enhanced context window capabilities, potentially reducing the administrative burden while maintaining or improving coding accuracy.
+Despite advances, significant challenges persist in automated medical coding systems.
+Clinical documents often exceed the standard context windows (e.g., 512 tokens) of many transformer models, causing potential information loss during truncation or requiring complex workarounds like the segment pooling employed by PLM-ICD [@huang-etal-2022-plm].
+Furthermore, the extreme class imbalance inherent in ICD coding hinders performance on rare but clinically significant codes.
+Although transformer-based approaches generally outperform older convolutional and recurrent architectures, these limitations highlight the need for architectures with natively expanded context windows, a capability offered by modern transformer variants like ModernBERT [@warner2024modernbert], which can process lengthy clinical documents more holistically.
 
-# Background
+# Methods {#sec-methods}
 
-@edin2024explainable recently demonstrated success in medical coding using a RoBERTa-based approach.
-Their work showed promising results but left room for improvement in both performance and contextual understanding.
-Other notable contributions in this field include [additional related work to be filled in].
+We followed the established data preparation pipeline developed by @cheng-etal-2023-mdace for processing the MIMIC-III and MIMIC-IV datasets [@johnson2023mimic], heavily relying on the open-source code provided by @edin2024explainable to match clinical visit notes with their associated diagnostic and procedure codes.
+Specifically, for the results compared against @edin2024explainable in @tbl-results, we utilized the MIMIC-III dataset [@johnson2016mimic], focusing on inpatient discharge summaries and their associated ICD-9 codes, processed according to the splits defined by @cheng-etal-2023-mdace.
+These notes originate from various clinicians across intensive care units at the Beth Israel Deaconess Medical Center.
+While previous approaches like PLM-ICD [@huang-etal-2022-plm] and its iterations employed specialized architectures for clinical text encoding, we diverged from this approach by adopting the more generic multi-label classification architecture recommended by ModernBERT [@warner2024modernbert].
+This decision simplified the architecture, facilitating experimentation and fully leveraging ModernBERT's 8192-token context window.
 
-ModernBERT [@warner2024modernbert] represents an advancement over previous BERT-like models, offering an enhanced context window of 8192 tokens compared to the 512 tokens in traditional models.
-This increased context capacity is particularly valuable for medical documents, which tend to be lengthy and contain important information distributed throughout the text.
+Training transformer models with such extended context windows presented significant computational challenges, particularly on consumer-grade hardware.
+To address memory constraints, we implemented a combination of gradient checkpointing during backpropagation and Flash Attention 2.0 [@dao2022flashattentionfastmemoryefficientexact].
+These optimizations enabled us to train our models on a standard consumer GPU without degrading performance.
+To facilitate direct comparison with existing approaches, we attempted to maintain the same evaluation metrics employed by PLM-ICD and subsequent works [@huang-etal-2022-plm; @edin2024explainable; @liu2023automated], including micro and macro-averaged F1 scores, precision, and recall metrics, which represent the standard evaluation framework in this field.
 
-## Using CSL
+# Results {#sec-results}
 
-# Methods
+Our proposed CLEM-ICD model, leveraging the ModernBERT architecture with an 8192-token context window, was evaluated on the MDACE dataset using the standard multi-label classification metrics.
+The best performing model achieved a Micro-averaged F1 score (Micro-F1) of 0.476, Micro-Precision of 0.680, and Micro-Recall of 0.366 on the test set.
+The Macro-averaged F1 score (Macro-F1) reached 0.038.
 
-This project utilizes the MIMIC-IV dataset, a large, freely available database comprising de-identified health data associated with hospital stays.
-The dataset includes detailed clinical notes, diagnostic codes, procedural information, and other health-related data from real hospital encounters.
+| Model                                 | Precision (%) | Recall (%) | Micro-F1 (%) |
+|---------------------------------------|---------------|------------|--------------|
+| AttInGrad (TM) [@edin2024explainable] | 40.2 $\pm$ 3.0  | 43.9 $\pm$ 4.8 | 41.9 $\pm$ 3.4 |
+| CLEM-ICD (Ours)                  | 68.0          | 36.6       | 47.6         |
 
-## System Architecture
+Table: Comparison of multi-label classification performance metrics between our CLEM-ICD model and the AttInGrad (TM) baseline from @edin2024explainable. {#tbl-results}
 
-Our approach replaces the RoBERTa model used in previous work [@edin2024explainable] with ModernBERT [@warner2024modernbert], leveraging its enhanced context window to better process lengthy clinical notes.
-We hypothesize that this will lead to improved code prediction by enabling the model to capture more comprehensive context from medical documents.
+These results demonstrate a notable improvement over recent benchmarks.
+For instance, comparing our Micro-F1 score of 47.6% to the 41.9% reported by @edin2024explainable for their AttInGrad model on the same MIMIC-III/MDACE benchmark dataset, our approach shows enhanced performance.
+A key difference is that the @edin2024explainable results are averaged over 10 runs, providing confidence intervals, whereas our CLEM-ICD results are based on a single training run due to computational constraints.
+Nevertheless, the CLEM-ICD model appears competitive with, and potentially surpasses, state-of-the-art methods like PLM-ICD [@huang-etal-2022-plm] under similar evaluation conditions, particularly benefiting from the extended context capacity.
+The low Macro-F1 score (3.8% on this dataset), however, aligns with observations in prior work, indicating persistent challenges in accurately classifying less frequent codes within the highly imbalanced ICD code distribution.
 
-## Implementation Details
+To provide a direct comparison on the widely used MIMIC-III benchmark dataset [@johnson2016mimic], we evaluate CLEM-ICD against the original PLM-ICD [@huang-etal-2022-plm] and subsequent improvements reported by @liu2023automated.
+The results are summarized in @tbl-mimic3-comparison.
+Our CLEM-ICD model demonstrates a strong Macro-F1 score, suggesting better performance on less frequent codes compared to prior work, although the Micro-F1 score is lower in this specific run.
 
-[To be completed with specific implementation details]
+| Model                     | Macro-F1 (%)   | Micro-F1 (%)   | P@5 (%)        | P@8 (%)        | P@15 (%)       |
+|---------------------------|----------------|----------------|----------------|----------------|----------------|
+| PLM-ICD [@huang-etal-2022-plm] | 10.4           | 59.8           | 84.4           | 77.1           | 61.3           |
+| BL-5 [@liu2023automated]   | 11.1 $\pm$ 0.1 | 60.7 $\pm$ 0.1 | 85.2 $\pm$ 0.2 | 78.0 $\pm$ 0.2 | 62.4 $\pm$ 0.1 |
+| CLEM-ICD (Ours)           | 16.5           | 54.6           | -              | -              | -              |
+: Comparison of results on the MIMIC-III full test set (%). CLEM-ICD results are from a single run. P@k metrics were not computed for CLEM-ICD in this run. {#tbl-mimic3-comparison}
 
-## Evaluation Approach
+Note that while Macro-F1 and Micro-F1 scores are directly comparable, the P@k metrics reported by @huang-etal-2022-plm and @liu2023automated differ from the overall micro-averaged precision (65.6%) and recall (46.8%) metrics recorded for our CLEM-ICD run.
 
-We evaluate our system using standard performance metrics: precision, recall, F1-score, and accuracy compared to gold-standard human coding, following the evaluation protocol established in previous work.
+::: {#fig-learning-curves-mdace layout-ncol="2" fig-cap="Learning curves for CLEM-ICD trained on MIMIC-III Inpatient Discharge Summaries (MDACE splits) for 10 epochs."}
 
-# Results and Discussion
+![Evaluation Recall (MDACE)](learning_curve_recall_mdace.png){#fig-recall-mdace}
 
-[This section will be completed after obtaining experimental results]
+![Evaluation Precision (MDACE)](learning_curve_precision_mdace.png){#fig-precision-mdace}
 
-# References {-}
+![Evaluation Macro-F1 Score (MDACE)](learning_curve_macro_f1_mdace.png){#fig-macro-f1-mdace}
+
+![Evaluation Micro-F1 Score (MDACE)](learning_curve_micro_f1_mdace.png){#fig-micro-f1-mdace}
+
+:::
+
+::: {#fig-learning-curves-full layout-ncol="2" fig-cap="Learning curves for CLEM-ICD trained on the full MIMIC-III dataset (PLM-ICD splits) for 20 epochs."}
+
+![Evaluation Recall (MIMIC-III Full)](learning_curve_recall_full.png){#fig-recall-full}
+
+![Evaluation Precision (MIMIC-III Full)](learning_curve_precision_full.png){#fig-precision-full}
+
+![Evaluation Macro-F1 Score (MIMIC-III Full)](learning_curve_macro_f1_full.png){#fig-macro-f1-full}
+
+![Evaluation Micro-F1 Score (MIMIC-III Full)](learning_curve_micro_f1_full.png){#fig-micro-f1-full}
+
+:::
+
+Observing the learning curves (@fig-learning-curves-mdace and @fig-learning-curves-full), the model trained on the full MIMIC-III dataset (@fig-learning-curves-full) appears to plateau relatively early in training, particularly for Micro-F1, while the model trained on the smaller MDACE split (@fig-learning-curves-mdace) shows more continued improvement later into the 10 epochs.
+Both runs exhibited signs of overfitting, suggesting that future training could benefit from implementing early stopping based on validation set performance.
+
+# Discussion
+
+## Architectural Considerations
+
+Our results underscore the advantage of leveraging transformer architectures with natively long context windows for processing lengthy clinical narratives in automated ICD coding.
+ModernBERT's 8192-token capacity allows CLEM-ICD to process entire documents holistically, potentially avoiding context fragmentation issues that can arise from the segment pooling techniques employed by prior work like PLM-ICD [@huang-etal-2022-plm] to handle shorter context limits.
+This architectural choice significantly simplifies the implementation, relying on standard Hugging Face library components (`AutoModelForSequenceClassification`) rather than requiring bespoke modules for segmentation or label-specific attention mechanisms, which were critical for PLM-ICD's performance.
+While this standard classification head might be considered less sophisticated than specialized approaches for extreme multi-label classification [@chang2019taming; @liu2023automated], the strong representational power of ModernBERT combined with its ability to access the full document context appears to compensate effectively for this task.
+
+## Base Model Choice and Efficiency
+
+The selection of ModernBERT-base [@warner2024modernbert] as the foundation model provides benefits beyond its extended context window.
+Incorporating architectural optimizations adapted from recent decoder-only models, ModernBERT-base (~138M parameters) is designed for improved computational efficiency, demonstrating enhanced inference speed compared to previous BERT-style encoders, particularly for variable-length inputs and long sequences due to its alternating attention mechanism [@warner2024modernbert].
+This increased efficiency holds potential for deployment in settings with limited computational resources or for enabling faster processing of large clinical datasets.
+However, the base model utilized here underwent general-domain pretraining (including text, code, and scientific literature) without specific adaptation to clinical or biomedical language.
+Consequently, further performance improvements might be realized through future domain-specific adaptation, either via continued pretraining on clinical corpora or fine-tuning biomedically-specialized ModernBERT variants.
+
+## Performance Analysis and Limitations
+
+The effectiveness of this approach is reflected in the competitive results.
+CLEM-ICD achieves a superior Micro-F1 on the MDACE benchmark compared to AttInGrad [@edin2024explainable] and demonstrates a markedly improved Macro-F1 score on the full MIMIC-III dataset compared to both PLM-ICD [@huang-etal-2022-plm] and the more recent BL-5 model [@liu2023automated].
+This strong Macro-F1 performance suggests that the model's access to the full, unsegmented context and potentially the broader knowledge within ModernBERT's pretraining data aids significantly in classifying less frequent codes, a persistent challenge in this domain.
+This capability is particularly relevant for developing coder assistance tools, as human coders often need more support with rare codes than common ones.
+
+However, the lower Micro-F1 score on the full MIMIC-III dataset compared to these benchmarks warrants consideration.
+This could be partly attributed to factors like variance from a single training run or differences in fine-tuning hyperparameters compared to the more established benchmarks.
+The learning curves also indicated potential overfitting (@fig-learning-curves-mdace, @fig-learning-curves-full), a known challenge when fine-tuning large language models [@dodge2020finetuning], suggesting that incorporating regularization techniques like early stopping is crucial.
+Furthermore, this study deliberately focused on classification performance.
+Consequently, it does not incorporate methods for enhancing model interpretability, such as identifying supporting text spans [@cheng-etal-2023-mdace] or visualizing attention patterns [@vaswani2017attention; @edin2024explainable], which remain important avenues for clinical NLP research.
+
+# Conclusion
+
+In this work, we presented CLEM-ICD, an automated ICD coding framework leveraging the ModernBERT architecture.
+By utilizing ModernBERT's native 8192-token context window, CLEM-ICD processes lengthy clinical documents effectively with a simplified architecture compared to previous methods like PLM-ICD that required complex segmentation or attention mechanisms.
+Our experiments demonstrate the potential of this approach, achieving competitive performance on the MDACE benchmark and notably strong Macro-F1 scores on the full MIMIC-III dataset, indicating improved classification of less frequent codes.
+This work highlights the promise of efficient, long-context encoder models for tackling the challenges of automated medical coding.
+To support reproducibility and further research, we release all code and model weights publicly.[^code]
+
+## Future Work
+
+Future research should focus on refining the fine-tuning process for large-context models in this domain.
+Implementing robust early stopping based on validation performance is a clear next step.
+Further investigation into techniques specifically addressing the long-tail distribution of ICD codes is warranted, including exploring specialized loss functions or hierarchical classification strategies [@liu2023automated].
+Evaluating the efficacy of parameter-efficient fine-tuning (PEFT) methods, such as LoRA [@hu2021lora], could also improve computational feasibility and potentially mitigate overfitting.
+The public release of our code and model weights aims to facilitate such investigations and contribute to the broader goal of developing robust, efficient, and potentially more interpretable automated medical coding systems.
+
+[^code]: Code and models available at \url{https://github.com/tylermarcuscross/explainable-medical-coding}
+
+# References

### 2025-04-15 article/article.tex
update

--- 2025-03-23 article/article.tex
@@
 \documentclass[
-  authoryear,
-  preprint]{elsarticle}
+]{article}
 
@@
     % xetex/luatex font selection
+    \setmainfont[]{Latin Modern Roman}
+  \setmathfont[]{Latin Modern Math}
 \fi
@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
+  \fi
+  % set entry spacing
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
+\usepackage{arxiv}
+\usepackage{orcidlink}
+\usepackage{amsmath}
+\usepackage[T1]{fontenc}
 \makeatletter
@@
 \makeatother
-\journal{Journal Name}
 
-\usepackage[]{natbib}
-\bibliographystyle{elsarticle-harv}
 \usepackage{bookmark}
@@
 \hypersetup{
-  pdftitle={CLEM-ICD - Clinical Language Encoding with ModernBERT},
-  pdfauthor={Tyler Cross},
-  pdfkeywords={ModernBERT, natural language
-processing, MIMIC-IV, ICD-10},
+  pdftitle={Clinical Language Encoding with ModernBERT},
+  pdfauthor={Tyler M Cross},
+  pdfkeywords={clinical natural language
+processing, ModernBERT, MIMIC-IV, MIMIC-III, MDACE, multi-label
+classification},
   colorlinks=true,
@@
 
-\setlength{\parindent}{6pt}
+\newcommand{\runninghead}{A Preprint }
+\renewcommand{\runninghead}{A Preprint }
+\title{Clinical Language Encoding with ModernBERT}
+\def\asep{\\\\\\ } % default: all authors on same column
+\author{\textbf{Tyler M Cross}~\orcidlink{0009-0003-3529-8222}\\Master's
+in Data Science\\University of California, Berkeley School of
+Information\\Berkeley,
+CA,\ 94720\\\href{mailto:tyler.cross@berkeley.edu}{tyler.cross@berkeley.edu}}
+\date{}
 \begin{document}
+\maketitle
+\begin{abstract}
+Medical coding---assigning standardized codes to clinical
+documentation---is a costly, labor-intensive process. Automating this
+task via multi-label classification faces challenges like long documents
+and extreme class imbalance. We present CLEM-ICD, replacing the
+RoBERTa/BERT base models common in prior frameworks with ModernBERT,
+leveraging its 8192-token context window to better process lengthy
+clinical notes. ModernBERT's optimized architecture (\textasciitilde138M
+parameters) enhances computational efficiency and simplifies
+implementation compared to prior approaches requiring complex
+modifications for shorter-context models. Evaluating on the MDACE
+dataset, CLEM-ICD achieves a Micro-F1 score of 47.6\%, outperforming
+recent results by Edin et al. (2024) (41.9\% Micro-F1). On the standard
+MIMIC-III benchmark, CLEM-ICD achieves a notably high Macro-F1 of
+16.5\%, significantly outperforming the current state-of-the-art on
+uncommon classes. CLEM-ICD demonstrates how modern architectural
+advancements, particularly large context windows applied to encoder-only
+models, can yield strong performance on complex tasks like automated
+medical coding. We release our code and models to foster further
+research.
+\end{abstract}
+{\bfseries \emph Keywords}
+\def\sep{\textbullet\ }
+clinical natural language
+processing \sep ModernBERT \sep MIMIC-IV \sep MIMIC-III \sep MDACE \sep 
+multi-label classification
 
-\begin{frontmatter}
-\title{CLEM-ICD - Clinical Language Encoding with
-ModernBERT \\\large{Using ModernBERT to Improve Automated ICD-10
-Classification} }
-\author[1]{Tyler Cross%
-\corref{cor1}%
-\fnref{fn1}}
- \ead{tyler.cross@berkeley.com}
+\section{Introduction}\label{sec-intro}
 
-\affiliation[1]{organization={University of California, Berkeley, School
-of Information},addressline={102 South
-Hall},city={Berkeley},postcode={94720},postcodesep={}}
+Medical coding, the assignment of standardized codes to clinical
+documentation, is an essential step for medical billing and
+reimbursement. The administrative processes surrounding billing and
+insurance-related activities, which heavily rely on accurate coding,
+represent a significant burden, consuming up to 25\% of healthcare
+expenditures in the United States (Kocher and Sahni 2011; Tseng et al.
+2018). Research has focused on attempts to automate coding by training
+multi-label classification models, with transformer-based architectures
+emerging as particularly effective (Huang, Tsai, and Chen 2022; Liu et
+al. 2024). The PLM-ICD framework (Huang, Tsai, and Chen 2022), utilizing
+BERT-based language models to encode clinical documents, established a
+strong benchmark for automated ICD coding by effectively balancing model
+complexity and predictive accuracy on the MIMIC-III dataset (A. E.
+Johnson et al. 2016).
 
-\cortext[cor1]{Corresponding author}
-\fntext[fn1]{Final project submission for the UC Berkeley Master of
-Information and Data Science program, DATASCI 266: Natural Language
-Processing with Deep Learning}
+Despite advances, significant challenges persist in automated medical
+coding systems. Clinical documents often exceed the standard context
+windows (e.g., 512 tokens) of many transformer models, causing potential
+information loss during truncation or requiring complex workarounds like
+the segment pooling employed by PLM-ICD (Huang, Tsai, and Chen 2022).
+Furthermore, the extreme class imbalance inherent in ICD coding hinders
+performance on rare but clinically significant codes. Although
+transformer-based approaches generally outperform older convolutional
+and recurrent architectures, these limitations highlight the need for
+architectures with natively expanded context windows, a capability
+offered by modern transformer variants like ModernBERT (Warner et al.
+2024), which can process lengthy clinical documents more holistically.
 
-\begin{abstract}
-Medical coding is a critical yet complex process that requires
-translating clinical documentation into standardized ICD-10 codes.
-Currently, healthcare providers rely on dedicated medical coders who
-manually review extensive clinical notes to assign appropriate codes
-from 150,000+ possible diagnostic and procedural codes. This project
-builds upon the work of \citet{edin2024explainable}{]} with two key
-innovations: (1) replacing RoBERTa with ModernBERT
-\citep{warner2024modernbert} as the encoder to potentially improve
-performance through its enhanced context window, and (2) adding a
-LLM-powered reasoning layer that provides natural language explanations
-for code predictions, bridging the gap between model predictions and
-human validation. Using the MIMIC-IV dataset, we implement and evaluate
-this approach using standard metrics (precision, recall, F1) and
-explanation quality.
-\end{abstract}
+\section{Methods}\label{sec-methods}
 
-\begin{keyword}
-    ModernBERT \sep natural language processing \sep MIMIC-IV \sep 
-    ICD-10
-\end{keyword}
-\end{frontmatter}
+We followed the established data preparation pipeline developed by Cheng
+et al. (2023) for processing the MIMIC-III and MIMIC-IV datasets (A. E.
+W. Johnson et al. 2023), heavily relying on the open-source code
+provided by Edin et al. (2024) to match clinical visit notes with their
+associated diagnostic and procedure codes. Specifically, for the results
+compared against Edin et al. (2024) in Table~\ref{tbl-results}, we
+utilized the MIMIC-III dataset (A. E. Johnson et al. 2016), focusing on
+inpatient discharge summaries and their associated ICD-9 codes,
+processed according to the splits defined by Cheng et al. (2023). These
+notes originate from various clinicians across intensive care units at
+the Beth Israel Deaconess Medical Center. While previous approaches like
+PLM-ICD (Huang, Tsai, and Chen 2022) and its iterations employed
+specialized architectures for clinical text encoding, we diverged from
+this approach by adopting the more generic multi-label classification
+architecture recommended by ModernBERT (Warner et al. 2024). This
+decision simplified the architecture, facilitating experimentation and
+fully leveraging ModernBERT's 8192-token context window.
 
-Medical coding is a critical yet complex process in healthcare
-administration. The translation of clinical documentation into
-standardized ICD-10 codes is essential for healthcare reimbursement,
-clinical research, and public health statistics. Currently, healthcare
-providers rely on dedicated medical coders who manually review extensive
-clinical notes to assign appropriate codes from over 150,000 possible
-diagnostic and procedural codes, creating significant administrative
-overhead.
+Training transformer models with such extended context windows presented
+significant computational challenges, particularly on consumer-grade
+hardware. To address memory constraints, we implemented a combination of
+gradient checkpointing during backpropagation and Flash Attention 2.0
+(Dao et al. 2022). These optimizations enabled us to train our models on
+a standard consumer GPU without degrading performance. To facilitate
+direct comparison with existing approaches, we attempted to maintain the
+same evaluation metrics employed by PLM-ICD and subsequent works (Huang,
+Tsai, and Chen 2022; Edin et al. 2024; Liu et al. 2024), including micro
+and macro-averaged F1 scores, precision, and recall metrics, which
+represent the standard evaluation framework in this field.
 
-For every \$1 of revenue collected by a hospital, approximately \$0.25
-is spent on administrative tasks necessary to collect it. Non-doctor
-workers outnumber doctors in healthcare roughly 16 to 1, with medical
-coding representing a significant portion of this burden.
+\section{Results}\label{sec-results}
 
-This project aims to improve automated medical coding by leveraging
-ModernBERT's enhanced context window capabilities, potentially reducing
-the administrative burden while maintaining or improving coding
-accuracy.
+Our proposed CLEM-ICD model, leveraging the ModernBERT architecture with
+an 8192-token context window, was evaluated on the MDACE dataset using
+the standard multi-label classification metrics. The best performing
+model achieved a Micro-averaged F1 score (Micro-F1) of 0.476,
+Micro-Precision of 0.680, and Micro-Recall of 0.366 on the test set. The
+Macro-averaged F1 score (Macro-F1) reached 0.038.
 
-\section{Background}\label{background}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.4875}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1875}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1500}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1750}}@{}}
+\caption{Comparison of multi-label classification performance metrics
+between our CLEM-ICD model and the AttInGrad (TM) baseline from Edin et
+al. (2024).}\label{tbl-results}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Model
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Precision (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Recall (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Micro-F1 (\%)
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Model
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Precision (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Recall (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Micro-F1 (\%)
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+AttInGrad (TM) (Edin et al. 2024) & 40.2 \(\pm\) 3.0 & 43.9 \(\pm\) 4.8
+& 41.9 \(\pm\) 3.4 \\
+CLEM-ICD (Ours) & 68.0 & 36.6 & 47.6 \\
+\end{longtable}
 
-\citet{edin2024explainable} recently demonstrated success in medical
-coding using a RoBERTa-based approach. Their work showed promising
-results but left room for improvement in both performance and contextual
-understanding. Other notable contributions in this field include
-{[}additional related work to be filled in{]}.
+These results demonstrate a notable improvement over recent benchmarks.
+For instance, comparing our Micro-F1 score of 47.6\% to the 41.9\%
+reported by Edin et al. (2024) for their AttInGrad model on the same
+MIMIC-III/MDACE benchmark dataset, our approach shows enhanced
+performance. A key difference is that the Edin et al. (2024) results are
+averaged over 10 runs, providing confidence intervals, whereas our
+CLEM-ICD results are based on a single training run due to computational
+constraints. Nevertheless, the CLEM-ICD model appears competitive with,
+and potentially surpasses, state-of-the-art methods like PLM-ICD (Huang,
+Tsai, and Chen 2022) under similar evaluation conditions, particularly
+benefiting from the extended context capacity. The low Macro-F1 score
+(3.8\% on this dataset), however, aligns with observations in prior
+work, indicating persistent challenges in accurately classifying less
+frequent codes within the highly imbalanced ICD code distribution.
 
-ModernBERT \citep{warner2024modernbert} represents an advancement over
-previous BERT-like models, offering an enhanced context window of 8192
-tokens compared to the 512 tokens in traditional models. This increased
-context capacity is particularly valuable for medical documents, which
-tend to be lengthy and contain important information distributed
-throughout the text.
+To provide a direct comparison on the widely used MIMIC-III benchmark
+dataset (A. E. Johnson et al. 2016), we evaluate CLEM-ICD against the
+original PLM-ICD (Huang, Tsai, and Chen 2022) and subsequent
+improvements reported by Liu et al. (2024). The results are summarized
+in Table~\ref{tbl-mimic3-comparison}. Our CLEM-ICD model demonstrates a
+strong Macro-F1 score, suggesting better performance on less frequent
+codes compared to prior work, although the Micro-F1 score is lower in
+this specific run.
 
-\subsection{Using CSL}\label{using-csl}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.2523}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1495}}@{}}
+\caption{Comparison of results on the MIMIC-III full test set (\%).
+CLEM-ICD results are from a single run. P@k metrics were not computed
+for CLEM-ICD in this run.}\label{tbl-mimic3-comparison}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Model
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Macro-F1 (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Micro-F1 (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+P@5 (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+P@8 (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+P@15 (\%)
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Model
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Macro-F1 (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Micro-F1 (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+P@5 (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+P@8 (\%)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+P@15 (\%)
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+PLM-ICD (Huang, Tsai, and Chen 2022) & 10.4 & 59.8 & 84.4 & 77.1 &
+61.3 \\
+BL-5 (Liu et al. 2024) & 11.1 \(\pm\) 0.1 & 60.7 \(\pm\) 0.1 & 85.2
+\(\pm\) 0.2 & 78.0 \(\pm\) 0.2 & 62.4 \(\pm\) 0.1 \\
+CLEM-ICD (Ours) & 16.5 & 54.6 & - & - & - \\
+\end{longtable}
 
-\section{Methods}\label{methods}
+Note that while Macro-F1 and Micro-F1 scores are directly comparable,
+the P@k metrics reported by Huang, Tsai, and Chen (2022) and Liu et al.
+(2024) differ from the overall micro-averaged precision (65.6\%) and
+recall (46.8\%) metrics recorded for our CLEM-ICD run.
 
-This project utilizes the MIMIC-IV dataset, a large, freely available
-database comprising de-identified health data associated with hospital
-stays. The dataset includes detailed clinical notes, diagnostic codes,
-procedural information, and other health-related data from real hospital
-encounters.
+\begin{figure}
 
-\subsection{System Architecture}\label{system-architecture}
+\begin{minipage}{0.50\linewidth}
 
-Our approach replaces the RoBERTa model used in previous work
-\citep{edin2024explainable} with ModernBERT
-\citep{warner2024modernbert}, leveraging its enhanced context window to
-better process lengthy clinical notes. We hypothesize that this will
-lead to improved code prediction by enabling the model to capture more
-comprehensive context from medical documents.
+\centering{
 
-\subsection{Implementation Details}\label{implementation-details}
+\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_recall_mdace.png}}
 
-{[}To be completed with specific implementation details{]}
+}
 
-\subsection{Evaluation Approach}\label{evaluation-approach}
+\subcaption{\label{fig-recall-mdace}Evaluation Recall (MDACE)}
 
-We evaluate our system using standard performance metrics: precision,
-recall, F1-score, and accuracy compared to gold-standard human coding,
-following the evaluation protocol established in previous work.
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
 
-\section{Results and Discussion}\label{results-and-discussion}
+\centering{
 
-{[}This section will be completed after obtaining experimental
-results{]}
+\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_precision_mdace.png}}
 
-\renewcommand\refname{References}
-  \bibliography{bibliography.bib}
+}
+
+\subcaption{\label{fig-precision-mdace}Evaluation Precision (MDACE)}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_macro_f1_mdace.png}}
+
+}
+
+\subcaption{\label{fig-macro-f1-mdace}Evaluation Macro-F1 Score (MDACE)}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_micro_f1_mdace.png}}
+
+}
+
+\subcaption{\label{fig-micro-f1-mdace}Evaluation Micro-F1 Score (MDACE)}
+
+\end{minipage}%
+
+\caption{\label{fig-learning-curves-mdace}Learning curves for CLEM-ICD
+trained on MIMIC-III Inpatient Discharge Summaries (MDACE splits) for 10
+epochs.}
+
+\end{figure}%
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_recall_full.png}}
+
+}
+
+\subcaption{\label{fig-recall-full}Evaluation Recall (MIMIC-III Full)}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_precision_full.png}}
+
+}
+
+\subcaption{\label{fig-precision-full}Evaluation Precision (MIMIC-III
+Full)}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_macro_f1_full.png}}
+
+}
+
+\subcaption{\label{fig-macro-f1-full}Evaluation Macro-F1 Score
+(MIMIC-III Full)}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{learning_curve_micro_f1_full.png}}
+
+}
+
+\subcaption{\label{fig-micro-f1-full}Evaluation Micro-F1 Score
+(MIMIC-III Full)}
+
+\end{minipage}%
+
+\caption{\label{fig-learning-curves-full}Learning curves for CLEM-ICD
+trained on the full MIMIC-III dataset (PLM-ICD splits) for 20 epochs.}
+
+\end{figure}%
+
+Observing the learning curves (Figure~\ref{fig-learning-curves-mdace}
+and Figure~\ref{fig-learning-curves-full}), the model trained on the
+full MIMIC-III dataset (Figure~\ref{fig-learning-curves-full}) appears
+to plateau relatively early in training, particularly for Micro-F1,
+while the model trained on the smaller MDACE split
+(Figure~\ref{fig-learning-curves-mdace}) shows more continued
+improvement later into the 10 epochs. Both runs exhibited signs of
+overfitting, suggesting that future training could benefit from
+implementing early stopping based on validation set performance.
+
+\section{Discussion}\label{discussion}
+
+\subsection{Architectural
+Considerations}\label{architectural-considerations}
+
+Our results underscore the advantage of leveraging transformer
+architectures with natively long context windows for processing lengthy
+clinical narratives in automated ICD coding. ModernBERT's 8192-token
+capacity allows CLEM-ICD to process entire documents holistically,
+potentially avoiding context fragmentation issues that can arise from
+the segment pooling techniques employed by prior work like PLM-ICD
+(Huang, Tsai, and Chen 2022) to handle shorter context limits. This
+architectural choice significantly simplifies the implementation,
+relying on standard Hugging Face library components
+(\texttt{AutoModelForSequenceClassification}) rather than requiring
+bespoke modules for segmentation or label-specific attention mechanisms,
+which were critical for PLM-ICD's performance. While this standard
+classification head might be considered less sophisticated than
+specialized approaches for extreme multi-label classification (Chang et
+al. 2019; Liu et al. 2024), the strong representational power of
+ModernBERT combined with its ability to access the full document context
+appears to compensate effectively for this task.
+
+\subsection{Base Model Choice and
+Efficiency}\label{base-model-choice-and-efficiency}
+
+The selection of ModernBERT-base (Warner et al. 2024) as the foundation
+model provides benefits beyond its extended context window.
+Incorporating architectural optimizations adapted from recent
+decoder-only models, ModernBERT-base (\textasciitilde138M parameters) is
+designed for improved computational efficiency, demonstrating enhanced
+inference speed compared to previous BERT-style encoders, particularly
+for variable-length inputs and long sequences due to its alternating
+attention mechanism (Warner et al. 2024). This increased efficiency
+holds potential for deployment in settings with limited computational
+resources or for enabling faster processing of large clinical datasets.
+However, the base model utilized here underwent general-domain
+pretraining (including text, code, and scientific literature) without
+specific adaptation to clinical or biomedical language. Consequently,
+further performance improvements might be realized through future
+domain-specific adaptation, either via continued pretraining on clinical
+corpora or fine-tuning biomedically-specialized ModernBERT variants.
+
+\subsection{Performance Analysis and
+Limitations}\label{performance-analysis-and-limitations}
+
+The effectiveness of this approach is reflected in the competitive
+results. CLEM-ICD achieves a superior Micro-F1 on the MDACE benchmark
+compared to AttInGrad (Edin et al. 2024) and demonstrates a markedly
+improved Macro-F1 score on the full MIMIC-III dataset compared to both
+PLM-ICD (Huang, Tsai, and Chen 2022) and the more recent BL-5 model (Liu
+et al. 2024). This strong Macro-F1 performance suggests that the model's
+access to the full, unsegmented context and potentially the broader
+knowledge within ModernBERT's pretraining data aids significantly in
+classifying less frequent codes, a persistent challenge in this domain.
+This capability is particularly relevant for developing coder assistance
+tools, as human coders often need more support with rare codes than
+common ones.
+
+However, the lower Micro-F1 score on the full MIMIC-III dataset compared
+to these benchmarks warrants consideration. This could be partly
+attributed to factors like variance from a single training run or
+differences in fine-tuning hyperparameters compared to the more
+established benchmarks. The learning curves also indicated potential
+overfitting (Figure~\ref{fig-learning-curves-mdace},
+Figure~\ref{fig-learning-curves-full}), a known challenge when
+fine-tuning large language models (Dodge et al. 2020), suggesting that
+incorporating regularization techniques like early stopping is crucial.
+Furthermore, this study deliberately focused on classification
+performance. Consequently, it does not incorporate methods for enhancing
+model interpretability, such as identifying supporting text spans (Cheng
+et al. 2023) or visualizing attention patterns (Vaswani et al. 2017;
+Edin et al. 2024), which remain important avenues for clinical NLP
+research.
+
+\section{Conclusion}\label{conclusion}
+
+In this work, we presented CLEM-ICD, an automated ICD coding framework
+leveraging the ModernBERT architecture. By utilizing ModernBERT's native
+8192-token context window, CLEM-ICD processes lengthy clinical documents
+effectively with a simplified architecture compared to previous methods
+like PLM-ICD that required complex segmentation or attention mechanisms.
+Our experiments demonstrate the potential of this approach, achieving
+competitive performance on the MDACE benchmark and notably strong
+Macro-F1 scores on the full MIMIC-III dataset, indicating improved
+classification of less frequent codes. This work highlights the promise
+of efficient, long-context encoder models for tackling the challenges of
+automated medical coding. To support reproducibility and further
+research, we release all code and model weights publicly.\footnote{Code
+  and models available at
+  \url{https://github.com/tylermarcuscross/explainable-medical-coding}}
+
+\subsection{Future Work}\label{future-work}
+
+Future research should focus on refining the fine-tuning process for
+large-context models in this domain. Implementing robust early stopping
+based on validation performance is a clear next step. Further
+investigation into techniques specifically addressing the long-tail
+distribution of ICD codes is warranted, including exploring specialized
+loss functions or hierarchical classification strategies (Liu et al.
+2024). Evaluating the efficacy of parameter-efficient fine-tuning (PEFT)
+methods, such as LoRA (Hu et al. 2021), could also improve computational
+feasibility and potentially mitigate overfitting. The public release of
+our code and model weights aims to facilitate such investigations and
+contribute to the broader goal of developing robust, efficient, and
+potentially more interpretable automated medical coding systems.
+
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-chang2019taming}
+Chang, Wei-Cheng, Hsiang-Fu Yu, Kai Zhong, Yiming Yang, and Inderjit S
+Dhillon. 2019. {``Taming Pretrained Transformers for Extreme Multi-Label
+Text Classification.''} In \emph{Proceedings of the 25th ACM SIGKDD
+International Conference on Knowledge Discovery \& Data Mining},
+318--28.
+
+\bibitem[\citeproctext]{ref-cheng-etal-2023-mdace}
+Cheng, Hua, Rana Jafari, April Russell, Russell Klopfer, Edmond Lu,
+Benjamin Striner, and Matthew Gormley. 2023. {``MDACE: MIMIC Documents
+Annotated with Code Evidence.''} In \emph{Proceedings of the 61st Annual
+Meeting of the Association for Computational Linguistics}, 7534--50.
+\url{https://doi.org/10.18653/v1/2023.acl-long.416}.
+
+\bibitem[\citeproctext]{ref-dao2022flashattentionfastmemoryefficientexact}
+Dao, Tri, Daniel Y. Fu, Stefano Ermon, Atri Rudra, and Christopher Ré.
+2022. {``FlashAttention: Fast and Memory-Efficient Exact Attention with
+IO-Awareness.''} \url{https://arxiv.org/abs/2205.14135}.
+
+\bibitem[\citeproctext]{ref-dodge2020finetuning}
+Dodge, Jesse, Gabriel Ilharco, Roy Schwartz, Ali Farhadi, Hannaneh
+Hajishirzi, and Noah Smith. 2020. {``Fine-Tuning Pretrained Language
+Models: Weight Initializations, Data Orders, and Early Stopping.''} In
+\emph{arXiv Preprint arXiv:2002.06305}.
+
+\bibitem[\citeproctext]{ref-edin2024explainable}
+Edin, Joakim, Maria Maistro, Lars Maaløe, Lasse Borgholt, Jakob D.
+Havtorn, and Tuukka Ruotsalo. 2024. {``An Unsupervised Approach to
+Achieve Supervised-Level Explainability in Healthcare Records.''}
+\emph{arXiv Preprint}. \url{https://arxiv.org/abs/2406.08958}.
+
+\bibitem[\citeproctext]{ref-hu2021lora}
+Hu, Edward J, Yelong Shen, Phillip Wallis, Zeyuan Allen-Zhu, Yuanzhi Li,
+Shean Wang, Lu Wang, and Weizhu Chen. 2021. {``LoRA: Low-Rank Adaptation
+of Large Language Models.''} In \emph{International Conference on
+Learning Representations}.
+
+\bibitem[\citeproctext]{ref-huang-etal-2022-plm}
+Huang, Chao-Wei, Shang-Chi Tsai, and Yun-Nung Chen. 2022. {``PLM-ICD:
+Automatic ICD Coding with Pretrained Language Models.''} In
+\emph{Proceedings of the 4th Clinical Natural Language Processing
+Workshop}, 10--20.
+\url{https://doi.org/10.18653/v1/2022.clinicalnlp-1.2}.
+
+\bibitem[\citeproctext]{ref-johnson2023mimic}
+Johnson, Alistair E. W., Lucas Bulgarelli, Lu Shen, Alvin Gayles, Ayad
+Shammout, Steven Horng, Tom J. Pollard, et al. 2023. {``MIMIC-IV, a
+Freely Accessible Electronic Health Record Dataset.''} \emph{Scientific
+Data} 10 (1): 1. \url{https://doi.org/10.1038/s41597-022-01899-x}.
+
+\bibitem[\citeproctext]{ref-johnson2016mimic}
+Johnson, Alistair EW, Tom J Pollard, Lu Shen, Li-Wei H Lehman, Mengling
+Feng, Mohammad Ghassemi, Benjamin Moody, Peter Szolovits, Leo Anthony
+Celi, and Roger G Mark. 2016. {``MIMIC-III, a Freely Accessible Critical
+Care Database.''} \emph{Scientific Data} 3 (1): 1--9.
+
+\bibitem[\citeproctext]{ref-kocher2011rethinking}
+Kocher, Robert, and Nikhil R. Sahni. 2011. {``Rethinking Health Care
+Labor.''} \emph{The New England Journal of Medicine} 365 (15): 1370--72.
+\url{https://doi.org/10.1056/NEJMp1109649}.
+
+\bibitem[\citeproctext]{ref-liu2023automated}
+Liu, Leibo, Oscar Perez-Concha, Anthony Nguyen, Vicki Bennett, and
+Louisa Jorm. 2024. {``Automated ICD Coding Using Extreme Multi-Label
+Long Text Transformer-Based Models.''}
+
+\bibitem[\citeproctext]{ref-tseng2018administrative}
+Tseng, Phillip, Robert S. Kaplan, Barak D. Richman, Mahek A. Shah, and
+Kevin A. Schulman. 2018. {``Administrative Costs Associated with
+Physician Billing and Insurance-Related Activities at an Academic Health
+Care System.''} \emph{JAMA} 319 (7): 691--97.
+\url{https://doi.org/10.1001/jama.2017.19148}.
+
+\bibitem[\citeproctext]{ref-vaswani2017attention}
+Vaswani, Ashish, Noam Shazeer, Niki Parmar, Jakob Uszkoreit, Llion
+Jones, Aidan N Gomez, Łukasz Kaiser, and Illia Polosukhin. 2017.
+{``Attention Is All You Need.''} In \emph{Advances in Neural Information
+Processing Systems}, 5998--6008.
+
+\bibitem[\citeproctext]{ref-warner2024modernbert}
+Warner, Benjamin, Antoine Chaffin, Benjamin Clavié, Orion Weller, Oskar
+Hallström, Said Taghadouini, Alexis Gallagher, et al. 2024. {``Smarter,
+Better, Faster, Longer: A Modern Bidirectional Encoder for Fast, Memory
+Efficient, and Long Context Finetuning and Inference.''} \emph{arXiv
+Preprint}. \url{https://arxiv.org/abs/2412.13663}.
+
+\end{CSLReferences}
 

### 2025-04-15 train_modernbert.py

@@ -234,7 +234,7 @@ def main(cfg: OmegaConf) -> None:
         num_labels=len(all_codes),
     )
 
-    # Enable gradient checkpointing to save memory (significant memory reduction)
+    # Enable gradient checkpointing to save memor
     model.gradient_checkpointing_enable()
     # Adjust epochs and other parameters for test mode
     num_epochs = 1 if test_mode else int(cfg.trainer.epochs)

### 2025-04-15 article/article.qmd
update

@@
   - ModernBERT
-  - MIMIC-IV
   - MIMIC-III

### 2025-04-15 article/article.tex
update

@@
   pdfkeywords={clinical natural language
-processing, ModernBERT, MIMIC-IV, MIMIC-III, MDACE, multi-label
-classification},
+processing, ModernBERT, MIMIC-III, MDACE, multi-label classification},
   colorlinks=true,
@@
 clinical natural language
-processing \sep ModernBERT \sep MIMIC-IV \sep MIMIC-III \sep MDACE \sep 
+processing \sep ModernBERT \sep MIMIC-III \sep MDACE \sep 
 multi-label classification

### 2025-04-15 train_modernbert.py
cleanup

@@ -226,7 +226,6 @@ def main(cfg: OmegaConf) -> None:
             "recall": recall,
         }
 
-    # Initialize model with gradient checkpointing for memory efficiency
     LOGGER.info(f"Loading model from {cfg.model.configs.model_path}")
     model = AutoModelForSequenceClassification.from_pretrained(
         cfg.model.configs.model_path,
@@ -234,7 +233,7 @@ def main(cfg: OmegaConf) -> None:
         num_labels=len(all_codes),
     )
 
-    # Enable gradient checkpointing to save memor
+    # Enable gradient checkpointing to save memory
     model.gradient_checkpointing_enable()
     # Adjust epochs and other parameters for test mode
     num_epochs = 1 if test_mode else int(cfg.trainer.epochs)
